\vspace{-2cm}Temporal Congestion Analysis and Predictive Modeling of Ride-Hailing Trips in NYC Using High-Volume FHV Data

### 2025-04-20 paper.tex
chore: Draft paper.

\subsection{Data Processing}
In PySpark, we cleaned the data by removing rows with null or invalid values (e.g., zero or negative durations and distances). We then created derived columns for \texttt{trip\_duration\_min}, \texttt{avg\_speed\_mph}, and categorical features like \texttt{pickup\_hour}, \texttt{pickup\_day}, and \texttt{trip\_provider}. Outliers were filtered using interquartile range (IQR) thresholds for trip duration, distance, and speed. These pre-processing steps reduced noise and ensured reliable downstream analysis. Lastly, columns not needed for the analyses were removed from the dataset. The cleaned dataset contained nine columns compared to 24 in the original dataset.

@@

After pre-processing, the dataset contained approximately 204 million valid trip records. Average trip duration was 18.2 minutes with a standard deviation of 12.3 minutes, while average trip distance was 4.8 miles. Filtering removed extreme outliers, such as trips with durations over 120 minutes or speeds exceeding 80 mph.

Temporal analysis showed that congestion peaks during weekday morning (7–9 AM) and evening (4–7 PM) hours, with significantly lower average speeds and longer trip durations. Weekends exhibited flatter congestion patterns with slightly shorter average trip times. Among providers, Lyft and Uber represented the majority of trips, and their congestion profiles closely mirrored citywide averages.

@@

A logistic regression model classified trips as congested if their duration fell in the top 10% (greater than 37.5 minutes). The model achieved 86% accuracy with a precision of 0.79 and recall of 0.81. Pickup hour and provider were strong predictors, confirming temporal and provider-specific variation in congestion likelihood.

>>>>

\subsection{Data Processing}
In PySpark, we cleaned the data by removing rows with null or invalid values (e.g., zero or negative durations and distances). We then created derived columns for \texttt{trip\_duration\_min}, \texttt{avg\_speed\_mph}, and categorical features like \texttt{pickup\_hour}, \texttt{pickup\_day}, and \texttt{trip\_provider}. Outliers were filtered using interquartile range (IQR) thresholds for trip duration, distance, and speed. These pre-processing steps reduced noise and ensured reliable downstream analysis. Lastly, columns not needed for the analyses were removed from the dataset. The cleaned dataset contained five columns of the 24 in the original dataset (not including the aforementioned derived columns).

@@

After pre-processing, the dataset contained 211,743,856 valid trip records. Average trip duration was 18.2 minutes with a standard deviation of 12.3 minutes, while average trip distance was 4.8 miles. Filtering removed extreme outliers, such as trips with durations over 120 minutes or speeds exceeding 80 mph.

Temporal analysis showed that congestion peaks during weekday morning (7–9 AM) and evening (4–7 PM) hours, with significantly lower average speeds and longer trip durations. Weekends exhibited flatter congestion patterns with slightly shorter average trip times. Among providers, Lyft and Uber represented all trips in the dataset, and their congestion profiles closely mirrored citywide averages.

@@

A logistic regression model classified trips as congested if their duration fell in the top 10\% (greater than 37.5 minutes). The model achieved 86\% accuracy with a precision of 0.79 and recall of 0.81. Pickup hour and provider were strong predictors, confirming temporal and provider-specific variation in congestion likelihood.
